---
title: "A quick tour of mclustCompositional"
date: "`r format(Sys.time(), '%d %b %Y')`"
author: "Luca Scrucca & Fedele Greco"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
    css: "vignette.css"
vignette: >
  %\VignetteIndexEntry{Model-Based Clustering and Density Estimation for Compositional Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = "center", 
               out.width = "90%",
               fig.width = 6, fig.height = 5,
               dev.args = list(pointsize=10),
               par = TRUE, # needed for setting hook 
               collapse = TRUE, # collapse input & output code in chunks
               warning = FALSE)

knit_hooks$set(par = function(before, options, envir)
  { if(before && options$fig.show != "none") 
       par(family = "sans", mar=c(4.1,4.1,1.1,1.1), mgp=c(3,1,0), tcl=-0.5)
})

plot_simplex <- function(x, col = "black", pch = 20,
                         labels = colnames(x), grid = TRUE, ...)
{

  xname <- deparse(substitute(x))
  x <- as.matrix(x)
  d <- ncol(x)
  stopifnot(d == 3)
  x <- x / rowSums(x)
  if(is.null(labels)) labels <- paste0(xname, 1:d)

  h <- sqrt(3) / 2
  xy <- cbind(x[, 2] + 0.5 * x[, 3], h * x[, 3])


  plot(NA, xlim = c(-0.08, 1.08), ylim = c(-0.06, h + 0.10),
       asp = 1, axes = FALSE, xlab = "", ylab = "",
       frame.plot = FALSE, ...)

  polygon(c(0, 1, 0.5), c(0, 0, h), border = "grey30", lwd = 1.5)

  if(grid) 
  {
    vals <- seq(0.2, 0.8, by = 0.2)
    for(v in vals) 
    {
      lines(c(0.5 * (1 - v), 1 - v), c(h * (1 - v), 0), col = "grey90")
      lines(c(v, v + 0.5 * (1 - v)), c(0, h * (1 - v)), col = "grey90")
      lines(c(0.5 * v, 1 - 0.5 * v), c(h * v, h * v), col = "grey90")
    }
  }

  points(xy, col = col, pch = pch, ...)
  text(c(-0.04, 1.04, 0.5), c(-0.04, -0.04, h + 0.05),
       labels = labels, xpd = NA)

  invisible(xy)
}

set.seed(1) # for exact reproducibility
```

```{r}
library(mclustCompositional)
```

## Overview

`mclustCompositional` fits Gaussian mixture models to compositional data by
working in isometric log-ratio (ILR) coordinates. The workflow is:

1. transform positive compositions to unconstrained ILR coordinates;
2. fit Gaussian mixtures with `mclust`;
3. select the best model by BIC or ICL;
4. evaluate densities, posterior probabilities, and classifications on the
   original compositional scale.

The closure constant is arbitrary: multiplying a composition by a positive
constant does not change its ILR coordinates.

## Simulated compositions via ILR transform and its inverse

The helper `ilr_inverse()` is convenient for simulating compositions from
Gaussian groups in ILR space:

```{r}
y = rbind(matrix(rnorm(80, mean = c(-1.2, 0.0), sd = 0.35),
                 ncol = 2, byrow = TRUE),
          matrix(rnorm(80, mean = c( 1.0, 0.8), sd = 0.35), 
                 ncol = 2, byrow = TRUE))
x = ilr_inverse(y)
```

Note that each row of `x` is a three-part composition summing to one:

```{r closure}
head(cbind(x, Sum = rowSums(x)))
```

The compositional data can be transformed to ILR coordinates with `ilr_transform()`:
```{r}
y = ilr_transform(x)
```
which recovers the original coordinates used for simulation.

Plot of the compositions on the simplex and their ILR coordinates:
```{r, echo=-1, fig.width=8, fig.height=4}
par(mfrow = c(1, 2))
plot(y, pch = 20)
plot_simplex(x)
```

## Density estimation

Use `densityMclustCompositional()` when the fitted density is the primary
object of interest:

```{r density-fit}
dens = densityMclustCompositional(x)
summary(dens, parameters = TRUE)
```

Predictions can return mixture densities, component densities, or posterior
probabilities.

```{r predict}
newx = x[1:5, ]
predict(dens, newx, what = "dens")
predict(dens, newx, what = "z")
```

A fitted density can be visualized in ILR coordinates or directly on the
simplex. The simplex plot below uses a triangular grid and evaluates the
compositional density with `predict()`.

```{r, fig.width=8, fig.height=4}
par(mfrow = c(1, 2))
ygrid = expand.grid(ilr1 = seq(min(y[,1]), max(y[,1]), length.out = 100),
                    ilr2 = seq(min(y[,2]), max(y[,2]), length.out = 100))
mclust::surfacePlot(data = ygrid,
                    parameters = dens$parameters,
                    what = "density", 
                    type = "contour")

h = sqrt(3)/2
sx = seq(0, 1, length.out = 100)
sy = seq(0, h, length.out = 100)
sgrid = expand.grid(sx = sx, sy = sy)
xgrid = with(sgrid, cbind(1 - sx - sy/(2*h),
                          sx - sy/(2*h),
                          sy/h))
colnames(xgrid) = colnames(x)
inside = apply(xgrid > 0, 1, all)
dgrid = rep(NA_real_, nrow(xgrid))
dgrid[inside] = predict(dens, xgrid[inside, ], what = "dens")
dgrid = matrix(dgrid, length(sx), length(sy))
plot_simplex(x, pch = NA)
contour(sx, sy, dgrid, add = TRUE, drawlabels = FALSE, col = "grey30")
polygon(c(0, 1, 0.5), c(0, 0, h), border = "grey30", lwd = 1.5)
```

## Model-based clustering

`MclustCompositional()` is the main clustering function. The example below
tries one- and two-component Gaussian mixtures with a common full covariance
model in ILR space.

```{r fit}
fit = MclustCompositional(x, criterion = "ICL")
summary(fit)
```

The posterior probabilities are stored in `fit$z`; classification uncertainty is
stored in `fit$uncertainty`.
```{r posterior}
head(fit$z)
summary(fit$uncertainty)
```

Plot of clustering on the simplex and using the ILR coordinates:
```{r, echo=-1, fig.width=8, fig.height=4}
par(mfrow = c(1, 2))
cols = mclust.options("classPlotColors")
plot(y, pch = 20, col = cols[fit$classification])
plot_simplex(x, col = cols[fit$classification])
```

```{r}

```
